\clearpage
\addcontentsline{toc}{chapter}{References}
\bibliographystyle{UNSWthesis}

\begin{thebibliography}{999}

\begin{comment}
\bibitem
{BBG} Benzinger, H., Berkson, E. and Gillespie, T.A.,
Spectral families of projections, semigroups, and differential operators,
\textit{Tran. Amer. Math. Soc.} \textbf{275} (1983), 431--475.

\bibitem
{BG Fourier} Berkson, E. and Gillespie, T.A.,
Fourier series criteria for operator decomposability,
\textit{Integral Equations Operator Theory} \textbf{9} (1986), 767--789.

\bibitem
{BG Spectral} Berkson, E., and Gillespie, T.A.,
Spectral decompositions and harmonic analysis on UMD spaces,
\textit{Studia Mathematica} \textbf{112}(1) (1994), 12--49.

\bibitem
{BGM} Berkson, E. Gillespie, T.A. and Muhly, P.S.,
Abstract spectral decompositions guaranteed by the Hilbert transform,
\textit{Proc. London Math. Soc. (3)} \textbf{53} (1986), 489--517.

\bibitem
{Bourgain telaviv} 
Bourgain, J., {\em Martingale transforms and geometry of Banach spaces},
Israel seminar on geometrical aspects of functional analysis
(1983/84), XIV, 16 pp., Tel Aviv Univ., Tel Aviv, 1984.

\bibitem
{Bourgain} Bourgain, J.,
Some remarks on Banach spaces in which martingale difference sequences are
unconditional,
\textit{Ark. Mat.} \textbf{21} (1983), 163--168.

\bibitem
{Bourg} Bourgain, J.,
Vector-valued singular integrals and the $H^1$-BMO duality,
\textit{Probability Theory and Harmonic Analysis} (ed. Chao, J. A.) (1986), 
1--19.

\bibitem
{Burk3} Burkholder, D.L.,
A geometrical characterisation of Banach spaces in which martingale difference
sequences are unconditional,
\textit{Ann. Probability} \textbf{9} (1981), 997--1011.

\bibitem
{Burk2} Burkholder, D.L.,
A geometric condition that implies the existence of certain singular integrals
of Banach-space valued functions,
\textit{Proceedings of Conference on Harmonic Analysis in Honor of Antoni
Zygmund} (Chicago, Illinois, 1981), Wadsworth Publishers: Belmont,
1983, pp. 270--286.

\bibitem
{Burk4} Burkholder, D.L.,
Martingales and Fourier analysis in Banach spaces,
\textit{Lecture Notes in Mathematics} \textbf{1206} (1986), 61--108.

\bibitem
{Burk1} Burkholder, D.L.,
Martingale transforms,
\textit{Ann. Math. Statist.} \textbf{37} (1966), 1494--1504.

\bibitem
{Coifman} Coifman R.R., and Weiss, G.,
\textit{Transference methods in analysis},
Regional Conference Series in Mathematics 31,
Amer. Math. Soc., Providence, R.I., 1977.

\bibitem
{Con} Conway, J.,
\textit{A Course in Functional Analysis},
Springer-Verlag: New York, 1985.

\bibitem
{Dowson} Dowson, H.,
\textit{Spectral theory of linear operators},
London Math. Soc. Monographs, No. 12. Academic Press: New York, 1978.

\bibitem
{Doust} 
Doust, I., Norms of $0$-$1$ matrices in $\CC_p$,  pp 50-55, Proc.
Centre Math. Appl. Austral. Nat. Univ., 39, Austral. Nat. Univ.,
Canberra, 2000.

\bibitem
{DG} Doust, I.  and Gillespie, T. A.,
Schur multipliers on $\CC_p$ spaces,
in preparation.

\bibitem
{Dun} Dunford, N., and Schwartz, J.T.
\textit{Linear operators I: General theory},
Pure and Applied Mathematics 7. Interscience: New York, 1958.

\bibitem
{Gaudry} Edwards, R.E., and Gaudry, G.I.,
\textit{Littlewood--Paley and Multiplier Theory},
Springer-Verlag: Berlin, 1977.

\bibitem
{Katznelson}
Katznelson, Y.,
\textit{An introduction to harmonic analysis},
Second corrected edition, Dover: New York, 1976.

\bibitem
{Gohberg}
Gohberg, I.C., and Kre\u{\i}n, M. G.,
\textit{Introduction to the theory of linear nonselfadjoint operators},
Translations of Mathematical Monographs, Vol. 18, Amer. Math. Soc.,
Providence, R.I., 1969.

\bibitem
{Gowers} Gowers, W. T., and Maurey, B.,
The unconditional basic sequence problem,
\textit{J. Amer. Math. Soc.} \textbf{6} (1993), 851--874.

\bibitem
{Lind}
Lindenstrauss, J., and Tzafriri, L.,
\textit{Classical Banach Spaces I and II},
Springer: Berlin, 1996.

\bibitem
{Pedersen}
Pedersen, G. K.,
\textit{Analysis Now},
Springer: New York, 1989.

\bibitem
{Stein1}
Stein, E. M.,
The development of square functions in the work of A. Zygmund,
\textit{Bulletin (New Series) of the Amer. Math. Soc},
\textbf{6} (1982), 5--30.

\bibitem
{Stein}
Stein, E. M.,
\textit{Topics in Harmonic Analysis Related to the Littlewood--Paley Theory},
Princeton UP: Princeton, 1970.
\end{comment}

\end{thebibliography}
