---
title: "Monte Carlo Error Calculations"
author: "Victor Tsang"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    highlight: pygments
header_includes:
  - \newcommand{\Var}{\mathrm{Var}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  attr.source='.numberLines'
)
setwd("C:\\Users\\victo\\OneDrive - UNSW\\UNSW\\22Hons\\Project Code\\")
library(readxl)
library(purrr)
```

## Monte Carlo Error

\begin{align*}
    \hat{\theta}_q
      &\approx K - q^{-\frac{1}{n}} \left[\frac{1}{B}\sum_{b=1}^B \prod_{i=1}^n (K - m + e_{ib}) \right]^\frac{1}{n} \\
      &= K - q^{-\frac{1}{n}}(K-m) \left[\frac{1}{B}\sum_{b=1}^B \prod_{i=1}^n \left(1 + \frac{e_{ib}}{K-m} \right) \right]^\frac{1}{n} \\
    \therefore \hat{\theta}_q
      &= K - q^{-\frac{1}{n}}(K-m) \left[\frac{1}{B}\sum_{b=1}^B Y_b \right]^\frac{1}{n}, \quad \text{where $Y_b = \prod_{i=1}^n \left(1 + \frac{e_{ib}}{K-m} \right) $} \\
\end{align*}

Applying the Delta Method to $\bar{Y_B} = \frac{1}{B} \sum_{b=1}^B Y_b$ with $h(x) = x^\frac{1}{n}$ lets us find the variance of $\bar{Y_B}^{1/n}$

\begin{align*}
    \Var \left\{h(\bar{Y_B})\right\}
      &\approx \left( h'(\mathbb{E} \bar{Y_B}) \right)^2 \Var{\bar{Y_B}} \\
      &\approx \left( \frac{1}{n} \left[ \mathbb{E}\bar{Y_B} \right]^{\frac{1}{n} - 1} \right)^2 \Var{\bar{Y_B}} \\
    \therefore \Var \hat{\theta}_q
      &\approx q^{-2/n} \left( \frac{1}{n} \left[ \mathbb{E}\bar{Y_B} \right]^{\frac{1}{n} - 1} \right)^2 \Var{\bar{Y_B}}
\end{align*}

We assumed $e_{ib}$ are independent and normally distributed with zero mean. Therefore $\mathbb{E}\bar{Y_B} = \mathbb{E}Y_B = \prod_{i=1}^n \left( 1 + \frac{\mathbb{E}e_{ib}}{K-m}\right) = 1$. Now let $\phi = \Var Y_b$.

\[
  \therefore \Var \hat{\theta}_q \approx q^{-2/n} n^{-2} B^{-1} \phi
\]

We can now use this to find a value of $B$ (call this $\tilde{B}$) that will give us variance below a target (call this target $\psi$)

\begin{align*}
    \Var\hat{\theta}_q &\leq \psi \\
    q^{-2/n} n^{-2} \tilde{B}^{-1} \phi &\leq \psi \\
    \tilde{B} &\geq \frac{\phi}{q^{2/n} n^{2} \psi}
\end{align*}

### Example

```{r ImportData}
set.seed(2022)

# Set hyper parameters
alpha = 0.025
K = 18000

# Import mammoth data from fossildata.xlsx
fossil_dataframe = read_excel(
  path='fossildata.xlsx', 
  sheet="MammothPrimEBer", 
  range="M3:N36", 
  col_names=c("age", "sd"), 
  col_types=c('numeric', 'numeric')
)
fossil_dataframe = fossil_dataframe[fossil_dataframe$age<K,]

W = fossil_dataframe$age
m = min(W)
n = length(W)

list(K=K, m=m, n=n)
```

Next, we generate our $\epsilon$ using some arbitrary $B_0$:

```{r}
B0 = 100
sigma = fossil_dataframe$sd

epsilon.B0 = rnorm(n=B0, mean=0, sd = sigma[1])
for (i in 2:n) {
  epsilon.B0 = cbind(epsilon.B0, rnorm(n=B0, mean=0, sd=sigma[i]))
}
cbind(epsilon.B0[1:5, 1:2], rep("...", 5))
```

Now, we use those generated errors to get an initial estimate of $\phi_{B_0}$

```{r}
Y.B0 = apply(1 + epsilon.B0/(K - m), 1, prod)
phi = var(Y.B0)
phi
```

Suppose we want our variance to be less than 100. Then:

```{r findOptimalB}
q = 0.025
psi = 100

B.tilde = ceiling(phi / (q^(2/n) * n^2 * psi))
B.tilde
```

Plot the size of B as our desired variance changes:

```{r, echo=FALSE}
target.vars = seq(50, 1000, length.out=100)
Bs = ceiling(target.vars / (q^(2/n) * n^2 * psi))
plot(x=target.vars, y=Bs, type="l", ylab="B", xlab=expression(psi))
print(paste("min(B):", min(Bs)))
print(list(K=K, m=m, n=n, q=q, phi=phi))
```









