---
title: "Monte Carlo Sample Calculations"
author: "Victor Tsang (z5209633)"
date: "`r Sys.Date()`"
output:
  html_notebook:
    highlight: pygments
  html_document:
    df_print: paged
    theme: united
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::knit_hooks$set(timeit = local({
  now = NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res = difftime(Sys.time(), now)
      now <<- NULL
      # use options$label if you want the chunk label as well
      paste('Time for this code chunk:', as.character(res))
    }
  }})
)

library(tikzDevice)

options(
    tikzLatexPackages = c(
        "\\usepackage{amsmath,amssymb,amsfonts}",
        "\\usepackage{tikz}",
        "\\usepackage{standalone}",
        "\\usepackage[utf8]{inputenc}",
        "\\usetikzlibrary{decorations.pathreplacing}",
        "\\usetikzlibrary{positioning}",
        "\\usetikzlibrary{shapes.arrows,shapes.symbols}",
        "\\usetikzlibrary{shapes.callouts}"
    )
)

options(tikzMetricsDictionary="../tikzMetrics") # speeds tikz up
```

#### Setup

Source our helper functions and set our parameters.

```{r}
source("../src/minmi-functions.R")

set.seed(2022)

n = 30                             # Number of fossil samples in each dataset
theta.true = 10000                 # True extinction date
K = 20000                          # Upper bound for fossil ages
 
dating_error.mean = 0              # True mean of radiocarbon dating error
dating_error.sd = 272              # True Standard deviation
 
alpha = 0.05                       # 95% confidence interval

```

First, simulate some data:

```{r}
W = runif(n=n, min=theta.true, max=K) + rnorm(n=n, mean=dating_error.mean, sd=dating_error.sd)
m=min(W)
list(theta.true=theta.true, m=m)
```


## Asymptotic Variance Formula

$$
\mathrm{Var}(\hat\theta_q) \longrightarrow \frac{1}{B} \sigma^2_{\psi(\theta_q)} \left[ \frac{f(K-\theta_q)}{F(K-\theta_q)}\hat\psi(\theta_q) + \hat\psi'(\theta) \right]^{-2}
$$

```{r}
estimate_var.delta_method = function (theta_q.hat, u, K, m, eps.mean, eps.sigma) {
  # pdfs and CDF evaluations (for convenience)
  f_eps.K = dnorm(K-theta_q.hat, eps.mean, eps.sigma)
  f_eps.m = dnorm(m-theta_q.hat, eps.mean, eps.sigma)
  
  F_eps.K = pnorm(K-theta_q.hat, eps.mean, eps.sigma)
  F_eps.m = pnorm(m-theta_q.hat, eps.mean, eps.sigma)
  
  # Estimate sigma^2_var
  B = length(u)
  e = uniform_to_tnorm(u, eps.mean, eps.sigma, a=-Inf, b=m-theta_q.hat)
  sample_var.psi_hat = var((m-e-theta_q.hat)/(K-e-theta_q.hat))
  
  # Estimate \hat\psi and \hat\psi prime
  psi_hat = mean((m-e-theta_q.hat)/(K-e-theta_q.hat))
  psi_hat.prime = - mean((K-m)/(K-e-theta_q.hat)^2)
  
  var_estimate = 1/B * sample_var.psi_hat * ( f_eps.K/F_eps.K * psi_hat + psi_hat.prime )^(-2)
  
  return(var_estimate)
}

estimate_var.delta_method(10180.92, u=runif(500, 0, 1), K, m, dating_error.mean, dating_error.sd)
estimate_var.delta_method(8441.491, u=runif(500, 0, 1), K, m, dating_error.mean, dating_error.sd)
```

## Checks

### Sample Variance of $\hat\theta_q$

$$
q^{\frac{1}{n}} = 1 - \frac{F_\varepsilon(m - \hat{\theta}_q)}{F_\varepsilon(K - \hat{\theta}_q)} \hat\psi_B \quad ; \quad \hat\psi_B =  \frac{1}{B} \sum_{b=1}^B \frac{m-e_b-\hat\theta_q}{K-e_b-\hat{\theta}_q}
$$

```{r}
u.gold = runif(5000, min=0, max=1)
theta.hat.lower.gold = estimate_quantile.minmi(q=0.025, K, W, u.gold,
                                               dating_error.mean,
                                               dating_error.sd)
theta.hat.upper.gold = estimate_quantile.minmi(q=0.975, K, W, u.gold,
                                               dating_error.mean,
                                               dating_error.sd)

print(list(lower.gold=theta.hat.lower.gold,
           upper.gold=theta.hat.upper.gold,
           m=m,
           theta.true=theta.true))
```

Compute sample and expected variances for a set of values of $B$

```{r}
n = 30

B = unique(round(pracma::logseq(2, 2000, 40)))
num.B = length(B)

num.estimates = 250

sample_var.theta_q = list()
sample_var.theta_q$lower = rep(NA, num.B)
sample_var.theta_q$upper = rep(NA, num.B)

expctd_var.theta_q = list()
expctd_var.theta_q$lower = rep(NA, num.B)
expctd_var.theta_q$upper = rep(NA, num.B)

u = matrix(runif(num.estimates*B[num.B], min=0, max=1), ncol=B[num.B])

start_time = Sys.time()
for (i in 1:num.B) {
  theta.hat.lower = rep(NA, num.estimates)
  theta.hat.upper = rep(NA, num.estimates)
  var_delta.lower = rep(NA, num.estimates)
  var_delta.upper = rep(NA, num.estimates)
  # Generate 100 estimates of \theta_q
  for (j in 1:num.estimates) {
    mc.samples = u[j, 1:(B[i])]
    
    theta.hat.lower[j] = estimate_quantile.minmi(q=0.025, K, W, mc.samples,
                                                 dating_error.mean,
                                                 dating_error.sd)
    theta.hat.upper[j] = estimate_quantile.minmi(q=0.975, K, W, mc.samples,
                                                 dating_error.mean,
                                                 dating_error.sd)
  }
  sample_var.theta_q$lower[i] = var(theta.hat.lower)
  sample_var.theta_q$upper[i] = var(theta.hat.upper)
  expctd_var.theta_q$lower[i] = estimate_var.delta_method(theta.hat.lower.gold,
                                                          u=u[1,1:(B[i])], K, m,
                                                          dating_error.mean, dating_error.sd)
  expctd_var.theta_q$upper[i] = estimate_var.delta_method(theta.hat.upper.gold,
                                                          u=u[1,1:(B[i])], K, m,
                                                          dating_error.mean, dating_error.sd)
}

time_taken = Sys.time() - start_time
print(time_taken)

res.df = data.frame(
  B = B,
  sample_var.upper = sample_var.theta_q$upper,
  expctd_var.upper = expctd_var.theta_q$upper,
  sample_var.lower = sample_var.theta_q$lower,
  expctd_var.lower = expctd_var.theta_q$lower
)

res.df[sample(nrow(res.df), 5), ]

lapply(list(ratio_lower=res.df$sample_var.lower/res.df$expctd_var.lower), FUN=function(x) c(min=min(x), max=max(x)))
lapply(list(ratio_upper=res.df$sample_var.upper/res.df$expctd_var.upper), FUN=function(x) c(min=min(x), max=max(x)))
```

Our expected variance from the delta method is approximately:

* equal to the sample variance for $q=0.025$.
* a third of the sample variance for $q=0.975$

### Plotting Sample Variances against Expected Variances

```{r fig.width=8, fig.height=5, dev='tikz'}
options(scipen=9)

plot(B, sample_var.theta_q$lower, log="xy",
     type="b",
     main="Sample Var vs. Asymptotic Var (q=0.025)", col="blue",
     ylim=c(min(sample_var.theta_q$lower, expctd_var.theta_q$lower), max(sample_var.theta_q$lower, expctd_var.theta_q$lower)),
     ylab=NA)
lines(B, expctd_var.theta_q$lower, type="b", col="red")
legend("topright",
       legend=c("Sample Var", "Asymptotic"),
       lty=c(1,1),
       pch=c(1,1),
       col=c("blue","red"))

```

```{r fig.width=8, fig.height=5, dev='tikz'}
# tikz("../figures/minmi-asymptotics-upper.tex")

plot(B, sample_var.theta_q$upper, log="xy",
     type="b",
     main="Sample Var vs. Asymptotic Var (q=0.975)", col="blue",
     ylim=c(min(sample_var.theta_q$upper, expctd_var.theta_q$upper), max(sample_var.theta_q$upper, expctd_var.theta_q$upper)),
     ylab=NA)
lines(B, expctd_var.theta_q$upper, type="b", col="red")
legend("topright",
       legend=c("Sample Var", "Asymptotic"),
       lty=c(1,1),
       pch=c(1,1),
       col=c("blue","red"))

```


- The left-most plot is the expected and sample variances for $q = 0.025$. This is pretty close to correct.
- The middle plot is the expected and sample variances for $q = 0.975$. Our sample is larger than our expected by a constant factor of ~3
- The right-most plot is the expected and sample variances for $q=0.975$ using $\hat\theta_q = K - q^{-1/n} * (K-m)$ as per our no measurement error case, instead of `theta.hat.upper.gold`. Sample is larger than expected by a factor of ~1.6

```{r dev='tikz'}
# tikz("../figures/minmi-asymptotics-ratios.tex")
options(scipen=9)

plot(B, res.df$expctd_var.upper/res.df$sample_var.upper, col="black",
     type="b", main="Delta Method Variance : Sample Variances | q=0.975", log="x",
     ylab=NA, ylim=c(min(c(res.df$expctd_var.lower/res.df$sample_var.lower, res.df$expctd_var.upper/res.df$sample_var.upper)), max(c(res.df$expctd_var.lower/res.df$sample_var.lower, res.df$expctd_var.upper/res.df$sample_var.upper))))

# lines(B, res.df$expctd_var.lower/res.df$sample_var.lower, type="b", col="red", ylab=NA)

# lines(lowess(x=B, y=res.df$expctd_var.lower/res.df$sample_var.lower), lty=3, col="blue")
lines(lowess(x=B, y=res.df$expctd_var.upper/res.df$sample_var.upper), lty=3, col="black")
# abline(h=mean(res.df$expctd_var.lower/res.df$sample_var.lower), lty=3, col="blue")
# abline(h=mean((res.df$expctd_var.upper)/res.df$sample_var.upper), lty=3, col="red")

# mtext(round(mean(res.df$expctd_var.lower/res.df$sample_var.lower), 3), side=4, las=2, at=mean(res.df$expctd_var.lower/res.df$sample_var.lower), col="blue")
# mtext(round(mean(res.df$expctd_var.upper/res.df$sample_var.upper), 3), side=4, las=2, at=mean(res.df$expctd_var.upper/res.df$sample_var.upper), col="black")

# abline(h=1.05)
# abline(h=0.95)
legend("topright",
       legend=c("Ratio","Lowess Smooth"),
       lty=c(1,2),
       pch=c(1,NA),
       col=c("blue","blue"))

```

```{r}
B[abs(res.df$expctd_var.upper/res.df$sample_var.upper - 1) < 0.05]
```

