---
title: "Monte Carlo Sample Calculations"
author: "Victor Tsang (z5209633)"
date: "`r Sys.Date()`"
output:
  html_notebook:
    highlight: pygments
  html_document:
    df_print: paged
    theme: united
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::knit_hooks$set(timeit = local({
  now = NULL
  function(before, options) {
    if (before) {
      now <<- Sys.time()
    } else {
      res = difftime(Sys.time(), now)
      now <<- NULL
      # use options$label if you want the chunk label as well
      paste('Time for this code chunk:', as.character(res))
    }
  }})
)
```

#### Setup

Source our helper functions and set our parameters.

```{r}
source("minmi-functions.R")

set.seed(2022)

n = 30                             # Number of fossil samples in each dataset
theta.true = 10000                 # True extinction date
K = 20000                          # Upper bound for fossil ages
 
dating_error.mean = 0              # True mean of radiocarbon dating error
dating_error.sd = 272              # True Standard deviation
 
alpha = 0.05                       # 95% confidence interval

```

First, simulate some data:

```{r}
W = runif(n=n, min=theta.true, max=K) + rnorm(n=n, mean=dating_error.mean, sd=dating_error.sd)
m=min(W)
list(theta.true=theta.true, m=m)
```


## Asymptotic Variance Formula

\[
\mathrm{Var}(\hat\theta_q) \longrightarrow \sigma^2_{\psi(\theta_q)} \left[ \left( - \frac{f(m - \theta)}{F(m - \theta)} + \frac{f(K - \theta)}{F(K - \theta)} \right) \hat\psi(\theta) + \hat\psi^\prime(\theta) \right]^{-2}
\]

```{r}
estimate_var.delta_method = function (q, theta_q.hat, B, K, m, eps.mean, eps.sigma) {
  # pdfs and CDF evaluations (for convenience)
  f_eps.K = dnorm(K-theta_q.hat, eps.mean, eps.sigma)
  F_eps.K = pnorm(K-theta_q.hat, eps.mean, eps.sigma)
  f_eps.m = dnorm(m-theta_q.hat, eps.mean, eps.sigma)
  F_eps.m = pnorm(m-theta_q.hat, eps.mean, eps.sigma)
  
  # Monte carlo samples
  e = matrix(extraDistr::rtnorm(100*B, eps.mean, eps.sigma, a=-Inf, b=m-theta_q.hat), ncol=B)
  
  # Estimate var(\hat\psi)
  sample_var.psi_hat = var(apply((m-e-theta_q.hat)/(K-e-theta_q.hat), 1, mean))
  
  # Estimate \hat\psi and \hat\psi prime
  psi_hat = mean((m-e[1,]-theta_q.hat)/(K-e[1,]-theta_q.hat))
  psi_hat.prime = - mean((K-m)/(K-e[1,]-theta_q.hat)^2)
  
  var_estimate = sample_var.psi_hat * ( psi_hat*(f_eps.K/F_eps.K - f_eps.m/F_eps.m) + psi_hat.prime)^(-2)
  
  return(var_estimate)
}

estimate_var.delta_method(q=0.975, 10181.15,
                          B=500, K, m,
                          dating_error.mean, dating_error.sd)
```

## Checks

### Sample Variance of $\hat\theta_q$

```{r}
n = 30

B = unique(round(pracma::logseq(2, 2000, 40)))
num.B = length(B)

num.estimates = 100

sample_var.theta_q = list()
sample_var.theta_q$lower = rep(NA, num.B)
sample_var.theta_q$upper = rep(NA, num.B)

expctd_var.theta_q = list()
expctd_var.theta_q$lower = rep(NA, num.B)
expctd_var.theta_q$upper = rep(NA, num.B)

u = matrix(runif(num.estimates*B[num.B], min=0, max=1), ncol=B[num.B])

uniroot.interval = c(5000, 19500)
```

```{r}
start_time = Sys.time()
for (i in 1:num.B) {
  theta.hat.lower = rep(NA, num.estimates)
  theta.hat.upper = rep(NA, num.estimates)
  var_delta.lower = rep(NA, num.estimates)
  var_delta.upper = rep(NA, num.estimates)
  # Generate 100 estimates of \theta_q
  for (j in 1:num.estimates) {
    mc.samples = u[j, 1:(B[i])]
    theta.hat.lower[j] = estimate_quantile.minmi(q=0.025, K, W, mc.samples,
                                                 dating_error.mean,
                                                 dating_error.sd, uniroot.interval)
    theta.hat.upper[j] = estimate_quantile.minmi(q=0.975, K, W, mc.samples,
                                                 dating_error.mean,
                                                 dating_error.sd, uniroot.interval)
    var_delta.lower[j] = estimate_var.delta_method(q=0.025, theta.hat.lower[j],
                                                   B[i], K, m,
                                                   dating_error.mean, dating_error.sd)
    var_delta.upper[j] = estimate_var.delta_method(q=0.975, theta.hat.upper[j],
                                                   B[i], K, m,
                                                   dating_error.mean, dating_error.sd)
  }
  sample_var.theta_q$lower[i] = var(theta.hat.lower)
  sample_var.theta_q$upper[i] = var(theta.hat.upper)
  expctd_var.theta_q$lower[i] = mean(var_delta.lower)
  expctd_var.theta_q$upper[i] = mean(var_delta.upper)
}
time_taken = Sys.time() - start_time
print(time_taken)

res.df = data.frame(
  B = B,
  sample_var.upper = sample_var.theta_q$upper,
  expctd_var.upper = expctd_var.theta_q$upper,
  sample_var.lower = sample_var.theta_q$lower,
  expctd_var.lower = expctd_var.theta_q$lower
)

lapply(list(ratio_lower=res.df$expctd_var.lower/res.df$sample_var.lower), FUN=function(x) c(min=min(x), max=max(x)))
lapply(list(ratio_upper=res.df$expctd_var.upper/res.df$sample_var.upper), FUN=function(x) c(min=min(x), max=max(x)))
```

Our expected variance from the delta method is approximately:

* equal to the sample variance for $q=0.025$.
* ~0.3x the sample variance for $q=0.975$

```{r fig.width=12, fig.height=4}
options(scipen=9)

par(mfrow=c(1,2))

plot(B, sample_var.theta_q$lower, log="xy",
     type="b", main="Sample Var vs. E(Delta method var) | q=0.025", col="blue",
     ylim=c(min(sample_var.theta_q$lower, expctd_var.theta_q$lower), max(sample_var.theta_q$lower, expctd_var.theta_q$lower)), ylab=NA)
lines(B, expctd_var.theta_q$lower, type="b", col="red")
legend("topright",
       legend=c("E(Delta Method Var)", "Sample Var"),
       lty=c(1,1),
       pch=c(1,1),
       col=c("red", "blue"))

plot(B, sample_var.theta_q$upper, log="xy",
     type="b", main="Sample Var vs. E(Delta method var) | q=0.975", col="blue",
     ylim=c(min(sample_var.theta_q$upper, expctd_var.theta_q$upper), max(sample_var.theta_q$upper, expctd_var.theta_q$upper)), ylab=NA)
lines(B, expctd_var.theta_q$upper, type="b", col="red")
legend("topright",
       legend=c("Sample Var", "E(Delta Method Var)"),
       lty=c(1,1),
       pch=c(1,1),
       col=c("blue","red"))
```


```{r fig.width=12, fig.height=4}
options(scipen=9)

par(mfrow=c(1,2))

plot(B, sample_var.theta_q$lower, log="x",
     type="b", main="Sample Var vs. E(Delta method var) | q=0.025", col="blue",
     ylim=c(min(sample_var.theta_q$lower, expctd_var.theta_q$lower), max(sample_var.theta_q$lower, expctd_var.theta_q$lower)), ylab=NA)
lines(B, expctd_var.theta_q$lower, type="b", col="red")
legend("topright",
       legend=c("E(Delta Method Var)", "Sample Var"),
       lty=c(1,1),
       pch=c(1,1),
       col=c("red", "blue"))

plot(B, sample_var.theta_q$upper, log="x",
     type="b", main="Sample Var vs. E(Delta method var) | q=0.975", col="blue",
     ylim=c(min(sample_var.theta_q$upper, expctd_var.theta_q$upper), max(sample_var.theta_q$upper, expctd_var.theta_q$upper)), ylab=NA)
lines(B, expctd_var.theta_q$upper, type="b", col="red")
legend("topright",
       legend=c("E(Delta Method Var)", "Sample Var"),
       lty=c(1,1),
       pch=c(1,1),
       col=c("red", "blue"))
```