---
title: "Monte Carlo Estimation Simulation Tests"
author: "Victor Tsang"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    highlight: pygments
    toc: true
    toc_float: true
    toc_collapsed: false
    toc_smooth_scroll: false
    toc_depth: 5
header_includes:
  - \newcommand{\Var}{\mathrm{Var}}
  - \newcommand{\E}{\mathbb{E}}
---

```{=html}
<style>
body
  { counter-reset: source-line 0; }
pre.numberSource code
  { counter-reset: none; }
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  attr.source='.numberLines',
  class.source = "numberLines lineAnchors",
  root.dir='C:/Users/victo/Documents/c-git-repos/HonoursThesis2022',
  error=TRUE
)
```

```{r}
source("helpers.R")
```


## Example: No measurement error

We want to see that for a 95% confidence interval ($\alpha = 0.05$), our coverage probability should be approximately 95%

```{r}
set.seed(2022)

nSim = 1000
alpha = 0.05
n = 30
theta = 12500
K = 18000

sim.no_error = simulateConfidenceIntervals(nSim, alpha, n, theta, K)

mean(sim.no_error$containsTheta)
```

```{r example1_plot}
plot.simCIs(sim.no_error[1:100, ], theta)
```

Our coverage is decent, although it does look like our intervals tend to underestimate $\theta$ fairly often.

## Example: Measurement error

We will keep the same parameters as above, but now we introduce measurement error.

Let's assume our measurement errors are normally distributed with zero-mean, and set the standard deviation to the **average standard deviation from some real data**

We will use $B = 10000$ for our monte carlo step.

```{r}
library(readxl)

# Import mammoth data from fossildata.xlsx
fossil_dataframe = read_excel(
  path='../data/fossildata.xlsx', 
  sheet="MammothPrimEBer", 
  range="M3:N36", 
  col_names=c("age", "sd"), 
  col_types=c('numeric', 'numeric')
)
fossil_dataframe = fossil_dataframe[fossil_dataframe$age<K,]
```


```{r example2_setup}
set.seed(2022)

B = 10000
eps.mean = 0
eps.sigma = mean(fossil_dataframe$sd)
uniroot.interval=c(10000, 16000)

sim.error = simulateConfidenceIntervals(nSim, alpha, n, theta, K, B, eps.mean, eps.sigma, uniroot.interval)
plot.simCIs(sim.error[1:100, ], theta)
```

```{r}
mean(sim.error$containsTheta)
```


